Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-08-09 manuscript/main.qmd
replies to reviewer comments, sensitivity ecological proxies

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the potential, underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Keating-Bitoni2011; @Bijl2009], lack. We test the robustness of this method using down-sampled, simulated temperature gradients.

@@

$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e. the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate, $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

<!-- Add justification of logistic function -->

>>>>

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitoni2011], lack. We test the robustness of this method using down-sampled, simulated temperature gradients.

@@

$$\nu_{j} = A~+\frac{K-A}{1+e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e. the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate, $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

We use this generalised logistic function because it can follow the equatorial and polar asymptotes observed in the modern, latitudinal SST gradient, but can also accommodate a variety of other shapes, while consisting of only four shape parameters. This flexibility is primarily achieved by shifting the location of the curve along the latitudinal axis by varying $M$, and by altering the steepness of the curve by varying $B$. For example, one limb of a second-order polynomial as in @Bijl2009 can be approximated by increasing $M$ towards high latitudes, and decreasing $B$ to reduce the steepness of the curve.
